---
title: "STAT 433 - HW1"
author: "Junhyeok Park"
date: "2022-09-20"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include = FALSE}
library(tidyverse)
library(nycflights13)
library(lubridate)
options(dplyr.summarise.inform = FALSE)
```

## Homework / in class exercise

### Q1. How many flights have a missing `dep_time`? What other variables are missing? What might these rows represent?
```{r}
flights1 = flights %>% mutate(problem1 = case_when(
    is.na(dep_time) ~ 1,
    !is.na(dep_time) ~ 0
  )) %>% filter(problem1 == 1)

flights1 %>% 
  summarise(n = sum(problem1))
```
There are 8255 of flights that have a missing `dep_time`.

```{r}
flights1 %>%
  filter(problem1 == 1) %>% head()
```
If `dep_time` is a missing value, then usually the row also has missing values for `dep_delay`, `arr_time`, `arr_delay`, and `air_time`. From this information, we might guess that the rows that have missing value for `dep_time` indicates flight cancellation.

### Q2. Currently `dep_time` and `sched_dep_time` are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.
```{r}
flights2 = flights %>% 
  mutate(dep_time = (dep_time %/% 100) * 60 + dep_time %% 100,
         sched_dep_time = (sched_dep_time %/% 100) * 60 + sched_dep_time %% 100)

flights2 %>% head()
```

### Q3. Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of canceled flights related to the average delay? Use multiple dyplr operations, all on one line, concluding with `ggplot(aes(x= ,y=)) + geom_point()`.
```{r}
# create a column that indicates if the flight is cancelled
# 1 indicates flight cancellation
flights2 %>%
    mutate(cancelled = case_when(
      is.na(dep_time) ~ 1,
      !is.na(dep_time) ~ 0
    )) %>%
  group_by(year, month, day) %>%
  summarise(avg_arr_del = mean(arr_delay[arr_delay > 0], na.rm=TRUE),
            avg_dep_del =  mean(dep_delay[dep_delay > 0], na.rm=TRUE),
            prop_cancelled = mean(cancelled)) %>%
  ggplot(aes(y=prop_cancelled)) + 
  geom_point(aes(x= avg_arr_del, color='Avg Arrival Delay'), alpha=0.6) +
  geom_point(aes(x= avg_dep_del, color='Avg Departure Delay'), alpha=0.5) + 
  xlab("Average Delay in Minutes") + 
  ylab("Proportion of Cancalled Flights per Day") + theme_minimal() +
  labs(color='Delay')
```

Notice that there is a positive correlation between average delays and the number of cancelled flights per day, meaning that if certain days have higher average delay then the proportion of cancelled flights is likely to be high as well.


