---
title: "STAT 433 - Exam1 Review"
author: "Junhyeok Park"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r include=FALSE}
library(tidyverse)
library(ggplot2)
library(nycflights13)
library(lubridate)
```


## Question 1

> Make a table that describes each plane. It should have a column for tailnum, another column for average arrival delay, and another for the year the plane was manufactured.

```{r}
flights %>%
  group_by(tailnum) %>%
  summarise(avg_arr_delay = mean(arr_delay, na.rm=T)) %>%
  left_join(planes %>% select(tailnum, year), by = "tailnum") %>%
  arrange(desc(avg_arr_delay))
```

## Question 2

> Make a table where each row is a day of the year. The first column is the date. The 2:4 columns give the number of (scheduled) departures from EWR, LGA, and JFK.

```{r}
flights %>%
  mutate(year_day = yday(time_hour)) %>%
  group_by(year_day, origin) %>%
  summarise(dep_num = n()) %>%
  pivot_wider(names_from = origin, values_from = dep_num)
```

## Question 3

> Make a table where each row is a day of the year. Each destination airport is a column. The elements (day x destination) give the number of flights to that destination. What should NA’s be?

All missing values should be 0 because in this case the implicit `NA`'s indicate that there was no flight to the airport on that day of the year.

```{r}
q3 = flights %>%
    mutate(year_day = yday(time_hour)) %>%
    group_by(year_day, dest) %>%
    summarise(num_dep = n()) %>%
    pivot_wider(names_from = dest, values_from = num_dep)

# replace the missing values with 0
q3[is.na(q3)] = 0
  
q3 %>% head()
```

## Question 4 

> Of the planes built before 1980, what is the average flight distance?

```{r}
flights %>% 
  select(tailnum, distance) %>% 
  left_join(planes %>% filter(year < 1980), by = "tailnum") %>%
  summarise(avg_flight_dist = mean(distance, na.rm=T))
```

## Question 5

> Compute the average arrival delay for each destination airport. Plot this against the airports altitude.

```{r}
flights %>%
  group_by(dest) %>%
  summarise(avg_arr_delay = mean(arr_delay, na.rm=T)) %>% 
  left_join(airports %>% select(faa, alt), by = c("dest" = "faa")) %>%
  ggplot(aes(x=alt, y=avg_arr_delay)) + 
  geom_point() +
  geom_smooth(method = "lm") + 
  xlab("Destination Airport Altitude") + 
  ylab("Average Arrival Delay") +
  ggtitle("The Average Arrival Delay vs. Destination Airport Altitude")
  
```

## Question 6

> Find the average arr_delay, by hour of the day, at EWR. Make a ggplot to show this.

```{r}
flights %>%
  filter(origin == "EWR") %>%
  mutate(day_hour = hour(time_hour)) %>%
  group_by(day_hour) %>%
  summarise(avg_arr_delay = mean(arr_delay, na.rm=T)) %>%
  ggplot(aes(x=day_hour, y=avg_arr_delay)) + 
  geom_line() + 
  xlab("Hour of the Day") + 
  ylab("Average Arrival Delay") + 
  ggtitle("Average Arrival Delay for the Hour of the Day")
```

## Question 7

> Find the average arr_delay, by hour of the day, for each of the 3 origin airports. Make a faceted ggplot, one facet for each origin airport.

```{r}
flights %>% 
  mutate(day_hour = hour(time_hour)) %>%
  group_by(origin, day_hour) %>%
  summarise(avg_arr_delay = mean(arr_delay, na.rm=T)) %>%
  ggplot(aes(x=day_hour, y=avg_arr_delay)) +
  geom_line() + 
  facet_wrap(~origin)
  
```


## Question 8

> Make a ggplot where the x axis is temperature and the y axis is delay in February. If you've done that, then facet by hour of the day and don't plot points. Instead, just plot a smooth line.

```{r}
flights %>%
  filter(month==2 & !is.na(arr_delay)) %>%
  left_join(weather %>% filter(month==2 & !is.na(temp)), by = c("origin", "year", "month", "day", "hour")) %>%
  ggplot(aes(x=temp, y=arr_delay)) +
  geom_smooth() + 
  facet_wrap(~hour) + 
  xlab("Temperature") + 
  ylab("Arrival Delay") + 
  ggtitle("Arrival Delay vs. Temperature of the Origin Airport for Each Departure Hour")
```


