---
title: "STAT 433 - HW1"
author: "Junhyeok Park"
date: "2022-09-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(nycflights13)
library(lubridate)
```

## Homework / in class exercise
1. How many flights have a missing dep_time? What other variables are missing? What might these rows represent?
```{r}
flights %>% count(is.na(dep_time))
```
There are 8255 of flights that have a missing `dep_time`.

2. Currently dep_time and sched_dep_time are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.
```{r}
flights2 = flights %>% 
  mutate(dep_time = (dep_time %/% 100) * 60 + dep_time %% 100,
         sched_dep_time = (sched_dep_time %/% 100) * 60 + sched_dep_time %% 100)
```

3. Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of canceled flights related to the average delay? Use multiple dyplr operations, all on one line, concluding with `ggplot(aes(x= ,y=)) + geom_point()`.
```{r}
# create a column that indicates if the flight is cancelled
# 1 indicates flight cancellation
flights2 %>%
    mutate(cancelled = case_when(
      is.na(dep_time) ~ 1,
      !is.na(dep_time) ~ 0
    )) %>%
  group_by(year, month, day) %>%
  summarise(avg_arr_del = mean(arr_delay[arr_delay > 0], na.rm=TRUE),
            avg_dep_del = mean(dep_delay[dep_delay > 0], na.rm=TRUE),
            sum_cancelled = sum(cancelled)) %>%
  ggplot(aes(x= avg_arr_del, y=sum_cancelled)) + 
  geom_point() +
  xlab("Average Arrival Delay in Minutes") + 
  ylab("# of Cancalled Flights per Day") + theme_minimal()
```




