---
title: "STAT 433 - HW2"
author: "Junhyeok Park"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(dplyr.summarise.inform = FALSE)
options(warn=-1)
```

```{r include = FALSE}
library(tidyverse)
library(nycflights13)
library(lubridate)
library(ggrepel)
library(ggspectra)
library(ggfortify)
library(ggpmisc)
```

## Question

> What time of day should you fly if you want to avoid delays as much as possible? Does this choice depend on anything? Season? Weather? Airport? Airline? Find three patterns (“null results” are ok!). Write your results into Rmarkdown. Include a short introduction that summarizes the three results. Then, have a section for each finding. Support each finding with data summaries and visualizations. Include your code when necessary. This shouldn’t be long, but it might take some time to find the things you want to talk about and lay them out in an orderly way.

## Motivation

In this short report, I analyzed the `flights` data set to see if there's any pattern among flights delay. The readers of this report should be able to see the pattern of the flights departure delay based on the three factors: departure day, weather, and the origin/airline of the flight. If 

## Pattern 1) Based on General Departure Day and Hour

It would be reasonable to think that certain weekday could be more hectic compared to the others for some reason. Based on the plot below, notice that the Friday has the largest average delay of 16.1 minutes and the Sunday has the lowest delay of about 7.6 minutes. The average percentage of flights cancelation also peaks on the Friday with 0.032% and 0.031% on the Thursday.

```{r echo=FALSE}
not_canceled = flights %>% 
  filter(!is.na(dep_time), !is.na(arr_time))

with_canceled = flights %>% 
  mutate(canceled = case_when(
    is.na(dep_delay) & is.na(arr_delay) ~ 1,
    !is.na(dep_delay) ~ 0
  ))
```


```{r echo=FALSE}
# by wday
not_canceled %>%
  mutate(wday = wday(time_hour)) %>% 
  group_by(wday) %>%
  summarise(mean_del = mean(dep_delay, na.rm=TRUE)) %>%
  arrange(mean_del) %>%
  ggplot(aes(x=as.factor(wday), y=mean_del)) +
  geom_col() +
  geom_text(aes(label = round(mean_del,1)), vjust = 1.5, colour = "white")+ 
  scale_x_discrete(breaks=c("1","2","3","4","5","6","7"),
        labels=c("Mon","Tue","Wed","Thur","Fri","Sat","Sun")) +
  xlab("Week Day") + 
  ylab("Mean Departure Delay in Minutes") +
  ggtitle("Mean Departure Delay per Week Day")

# mean cancelation per wday
with_canceled %>% 
  mutate(wday = wday(time_hour, label=TRUE)) %>% 
  group_by(wday) %>%
  summarise(mean_can = mean(canceled)) %>% 
  ggplot(aes(x=as.factor(wday), y=mean_can)) +
  geom_col() +
  geom_text(aes(label = round(mean_can,4)), vjust = 1.5, colour = "white")+ 
  xlab("Week Day") + 
  ylab("Percentage of Cancelation") +
  ggtitle("Average Percentage of Flights Cancelation per Week Day")

```

The distributions of the average delay based on hours / weekdays are as follows.

```{r echo=FALSE}
not_canceled %>% 
  group_by(hour) %>%
  summarise(mean_del = mean(dep_delay, na.rm=TRUE)) %>%
  ggplot(aes(x=as.factor(hour), mean_del)) +
  geom_col(stat='identity') +
  geom_text(aes(label = round(mean_del,1)), vjust = 1.2, colour = "white", size=2.5) + 
  xlab("Departure Hour") + 
  ylab("Avg Departure Delay in Minutes") +
  ggtitle("Average Departure Delay per Hour")

not_canceled %>% 
  mutate(wday = wday(time_hour, label=TRUE)) %>%
  group_by(wday, hour) %>%
  summarise(mean_del = mean(dep_delay, na.rm=TRUE)) %>%
  ggplot(aes(x=as.factor(hour), y=mean_del)) +
  geom_col() +
  facet_grid(rows=vars(wday)) +
  xlab("Departure Hour") +
  ylab("Average Departure Delay") +
  ggtitle("Average Departure Delay per Weekday and Hour")
```

## Pattern 2) Weather

Over the year, following dates (dates with red dots - let's call these dates 'peak dates') have the average departure delay greater than 40 minutes. Notice these dates have the average precipitation that is almost 5 times greater than that of the other dates in inches. Also, the visibility is shorter in about 1.3 miles than that of the others. Interestingly, the average wind speed and pressure are actually lower on these 'peak' dates.

```{r echo=FALSE}
# line graph with peaks
df = flights %>%
  mutate(dep_date = make_date(year, month, day)) %>% 
  group_by(dep_date) %>%
  summarise(mean_del = mean(dep_delay, na.rm=T))

df %>%
  ggplot(aes(x=dep_date, y=mean_del)) +
  geom_line() + 
  geom_point(data = . %>% filter(mean_del>40), color = "red") + 
  ylim(-10, 90)

# save those dates with the average of departure delay greater than 40 minutes 
df_g40 = df %>%
    filter(mean_del > 40)
```

The dates with the red dots on them are as follows:

```{r}
# merge two different data sets
merged = flights %>%
    left_join(weather, by=c("year", "month", "day", "hour", 'origin'))
```

```{r}
# weather factors grouped by peak
peaks_dates = as.character(df_g40[["dep_date"]])

merged %>% 
  mutate(date = format(time_hour.x, "%Y-%m-%d"),
         is_peak = case_when(
           date %in% peaks_dates ~ "1",
           TRUE ~ "0"
         )) %>%
  group_by(is_peak) %>%
  summarise(avg_wind_speed = mean(wind_speed, na.rm=T),
            avg_precip = mean(precip, na.rm=T),
            avg_pressure = mean(pressure, na.rm=T),
            avg_visib = mean(visib, na.rm=T))
```

Notice that the percentage of flights cancelation is almost 6 times greater on the peak dates. 

```{r echo=FALSE}
# the average wind speed of canceled and not canceled flights
merged %>%
  mutate(canceled = case_when(
    is.na(dep_delay) & is.na(arr_delay) ~ 'canceled',
    !is.na(dep_delay) ~ 'not canceled'
  )) %>% 
  select(canceled, wind_speed) %>% 
  filter(!is.na(wind_speed)) %>% 
  group_by(canceled) %>%
  summarise(avg_wind_speed = mean(wind_speed))

# flight cancelation grouped by 40 min mark
df_40min = merged %>% 
  mutate(date = format(time_hour.x, "%Y-%m-%d"),
         is_peak = case_when(
           date %in% peaks_dates ~ "1",
           TRUE ~ "0"
         ))

df_40min %>% 
  mutate(canceled = case_when(
    is.na(dep_delay) & is.na(arr_delay) ~ 1,
    TRUE ~ 0
  )) %>%
  group_by(is_peak) %>%
  summarise(mean_can = mean(canceled)) %>%
  ggplot(aes(x=is_peak, y=mean_can,
             fill=factor(ifelse(is_peak==1,"Hlighted","Normal")))) +
  scale_fill_manual(name = "area", values=c("tomato1","grey50")) +
  geom_col(width=0.5, show.legend = FALSE) +
  geom_text(aes(label = paste(round(mean_can,4), "%"))
            , vjust = 1.5, colour = "white", size=4) +
  xlab("Not Peak / Peak Dates") + 
  ylab("Average Percentage of Flights Cancelation")
```

Let's look up what happened on those peak dates. The dates are as follows: 

```{r echo=FALSE}
peaks_dates
```

From the line graph with the red dots, we know that the two days with the worst delays are March 08 and July 22.

- March 8th, 2013 New York Weather - [Snow storm blankets New York City, leaving pretty scenes and a sloshy commute](https://www.nbcnews.com/news/photo/snow-storm-blankets-new-york-city-leaving-pretty-scenes-sloshy-flna1c8776255)

- July 22th, 2013, at the LaGuardia Airport, there was a crash. [Southwest Airlines Flight 345](https://en.wikipedia.org/wiki/Southwest_Airlines_Flight_345#:~:text=On%20July%2022%2C%202013%2C%20the,a%20result%20of%20the%20accident.&text=IATA%20flight%20No.&text=ICAO%20flight%20No.) So the delay was probably not due to the weather, but the accident.

## Pattern 3) Time and Day based on the Airport Origin

Now, let's investigate if there are certain airports that have bad delays compare to other airports.

First of all, the average flights delay and cancelation percentages for each airport in New York City as as follows:

```{r echo=FALSE}
# average delay per origin
flights %>%
  mutate(canceled = case_when(
    is.na(dep_delay) & is.na(arr_delay) ~ 1,
    !is.na(dep_delay) ~ 0
  )) %>% 
  filter(canceled == 0) %>%
  group_by(origin) %>%
  summarise(avg_delay = mean(dep_delay, na.rm=T)) %>%
  ggplot(aes(x=origin, y=avg_delay)) +
  geom_col(width = 0.5) +
  xlab("Origin") + 
  ylab("Average Delay in Minutes") + 
  ggtitle("Average Delay for Each Origin")

# average number of flights cancelation per origin
flights %>% 
  mutate(canceled = case_when(
    is.na(dep_delay) & is.na(arr_delay) ~ 1,
    !is.na(dep_delay) ~ 0
  )) %>% 
  group_by(origin) %>%
  summarise(avg_cancelation = mean(canceled, na.rm=T)) %>%
  ggplot(aes(x=origin, y=avg_cancelation)) + 
  geom_col(width=0.5) + 
  xlab("Origin") + 
  ylab("Average cancelation Percentage") + 
  ggtitle("Average cancelation Percentage for Each Origin")
```

Let's do the same analysis on the airlines. 

```{r echo=FALSE}
# per airlines - delays
flights %>%
  left_join(airlines, by='carrier') %>%
  group_by(name) %>%
  summarise(avg_delay = mean(dep_delay, na.rm=T)) %>%
  ggplot(aes(x=fct_rev(fct_reorder(name, avg_delay)), y=avg_delay)) + 
  geom_col(width=0.5) +
  theme(axis.text.x = element_text(angle = 45, hjust=1, size=5)) + 
  xlab("Airline") + 
  ylab("Average Delay in Minutes") + 
  ggtitle("Average Delay for Each Airline")
  
# per airlines - cancelation
flights %>%
  left_join(airlines, by='carrier') %>%
  mutate(canceled = case_when(
    is.na(dep_delay) & is.na(arr_delay) ~ 1,
    !is.na(dep_delay) ~ 0
  )) %>% 
  group_by(name) %>%
  summarise(avg_canceled = mean(canceled, na.rm=T)) %>%
  ggplot(aes(x=fct_rev(fct_reorder(name, avg_canceled)), y=avg_canceled)) + 
  geom_col(width=0.5) +
  theme(axis.text.x = element_text(angle = 45, hjust=1, size=6)) +
  xlab("Airline") + 
  ylab("Average cancelation Percentage") + 
  ggtitle("Average Cancelation Percentage for Each Airline")
```

These following lists show that the 'bad combination' of airports and airlines. First, we have a data with three airlines with the worst average flights delay for each airport.

```{r echo=FALSE}
# top 3 airlines with the worst delay for each airport
flights %>% 
  left_join(airlines, by='carrier') %>%
  group_by(origin, name) %>%
  summarise(avg_del = mean(dep_delay, na.rm=T)) %>%
  # arrange(desc(avg_del), .by_group = TRUE) %>%
  slice_max(order_by = avg_del, n=3)
```

Lastly, the following data shows the airlines with the worst average percentage of flights cancelation for each airport.

```{r echo=FALSE}
# top 3 airlines with the worst cancelation for each airport
flights %>%
  left_join(airlines, by='carrier') %>%
  mutate(canceled = case_when(
    is.na(dep_delay) & is.na(arr_delay) ~ 1,
    !is.na(dep_delay) ~ 0
  )) %>%
  group_by(origin, name) %>%
  summarise(avg_cancelation = mean(canceled, na.rm=T)) %>%
  slice_max(order_by = avg_cancelation, n=3)
```

## Conclusion

In this report, a preliminary analysis is done on the `nycflights` data set to distinguish the pattern of flights delay and cancelation percentage based on three factors: departure time, weather, and airport/airlines. The Friday has the worst delay and flights cancelation percentage. Also, try to avoid flights with the departure hour between 18:00 - 21:00. 