---
title: "STAT 433 - HW3"
author: "Junhyeok Park"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(dplyr.summarise.inform = FALSE)
options(warn=-1)
```

```{r include = FALSE}
library(tidyverse)
library(nycflights13)
library(lubridate)
library(ggplot2)
```

# Questions in r4ds, section 13.4.6, numbers 1,2,3

## Exercise 1

> Compute the average delay by destination, then join on the airports data frame so you can show the spatial distribution of delays.

```{r}
# the average delay by destination
flights %>%
  group_by(dest) %>%
  summarise(mean_del = mean(dep_delay, na.rm=T)) %>% 
  left_join(airports, by=c("dest"="faa")) %>%
  ggplot(aes(lon, lat)) +
    borders("state") +
    geom_point(aes(color=mean_del)) +
    coord_quickmap() + 
    xlab("Longitude") + 
    ylab("Latitude") + 
    ggtitle("Average Departure Delay per Destination") +
    labs(color="Average Delay")
```

## Exercise 2

> Add the location of the origin and destination (i.e. the lat and lon) to flights.

```{r}
flights %>%
  left_join(airports %>% select(faa, lat, lon), by = c("origin"="faa")) %>%
  left_join(airports %>% select(faa, lat, lon), by = c("dest"="faa")) %>%
  rename(
    "origin_lat" = "lat.x",
    "origin_lon" = "lon.x",
    "dest_lat" = "lat.y",
    "dest_lon" = "lon.y"
  ) %>%
  select(year:origin, origin_lat, origin_lon, dest, dest_lat, dest_lon, everything()) %>% 
  head()
```

## Exercise 3

> Is there a relationship between the age of a plane and its delays?

The first graph 'naively' plotted the scatter plot of the age of planes and the departure delay of the flight. In the first glance, there seems to be a negative linear relationship between them. However, it is not safe to make such conclusion mainly because it might be just the difference of the sizes: there are simply much more planes aged between 0 and 40 and, thus, they are more likely to have delays. Let's remove outliers - flights with delays greater than 500 minutes.

```{r}
flights %>%
  left_join(planes %>% mutate(plane_age = 2022 - year) %>% select(tailnum, plane_age),
            by = "tailnum") %>%
  # ignoring missing values
  filter(!is.na(dep_delay) & !is.na(plane_age)) %>%
  ggplot(aes(x=plane_age, y=dep_delay)) + 
  geom_point() + 
  xlab("Plane Age") + 
  ylab("Departure Delay (Min)") + 
  ggtitle("Correlation between the Age of Planes and Departure Delay")
```

Notice that there is not a strong correlation between the age of planes and delays from the plot below. Let's investigate further by creating a new column: age groups.

```{r}
flights %>%
  left_join(planes %>% mutate(plane_age = 2022 - year) %>% select(tailnum, plane_age),
            by = "tailnum") %>%
  filter(!is.na(dep_delay) & !is.na(plane_age) & dep_delay < 500) %>%
  ggplot(aes(x=plane_age, y=dep_delay)) + 
  geom_point(alpha=0.5) +
  geom_smooth(method="lm") +
  xlab("Plane Age") + 
  ylab("Departure Delay (Min)") + 
  ggtitle("Correlation between the Age of Planes and Departure Delay - Outliers Removed")
```
Notice the variance of delays for each plane age group differs significantly. To take account for this difference, let's compare the mean delay for each group. 

```{r}
# create a data frame with planes age groups
plane_age_group = planes %>%
    mutate(plane_age = 2022 - year,
           age_group = case_when(
             plane_age >= 0 & plane_age <= 10 ~ "0-10",
             plane_age >= 11 & plane_age <= 20 ~ "11-20",
             plane_age >= 21 & plane_age <= 30 ~ "21-30",
             plane_age >= 31 & plane_age <= 40 ~ "31-40",
             plane_age >= 41 & plane_age <= 50 ~ "41-50",
             plane_age >= 51 & plane_age <= 60 ~ "51-60",
             plane_age >= 61 ~ "60"
             )) %>%
    select(tailnum, age_group) 

# join the data frames
flights %>% 
  left_join(plane_age_group, by = "tailnum") %>% 
  filter(!is.na(dep_delay) & !is.na(age_group) & dep_delay < 500) %>%
  ggplot(aes(x=age_group, y=dep_delay)) + 
  geom_point(alpha=0.5) + 
  xlab("Plane Age Group") + 
  ylab("Average Departure Delay (Min)") + 
  ggtitle("The Age Group vs. Departure Delay (Outliers Removed)")
```

The average delay of the planes with the ages bewteen 11-20 is the highest (14.3 min)

```{r include=FALSE}
# from http://www.sthda.com/english/wiki/ggplot2-error-bars-quick-start-guide-r-software-and-data-visualization
data_summary <- function(data, varname, groupnames){
  require(plyr)
  summary_func <- function(x, col){
    c(mean = mean(x[[col]], na.rm=TRUE),
      sd = sd(x[[col]], na.rm=TRUE))
  }
  data_sum<-ddply(data, groupnames, .fun=summary_func,
                  varname)
  data_sum <- rename(data_sum, c("mean" = varname))
 return(data_sum)
}

# create the df with age group
flights_age_group = flights %>% 
  left_join(plane_age_group, by = "tailnum") %>% 
  filter(!is.na(dep_delay) & !is.na(age_group) & dep_delay < 500)

flights_summary = data_summary(flights_age_group, varname="dep_delay",groupnames=c("age_group"))
```

Because of the large standard deviation of the departure delay for each plane age group, it is hard to draw a conclusion if there's any significant correlation between the age and the average of departure delay. Only with the average values, the planes with age between 11-20 has the highest mean delay with 14.3 minutes and those with age between 41-50 has the lowest average delay with 8.1 minutes.

```{r}
# plotting bar graph with error lines
flights_summary %>%
  ggplot(aes(x=age_group,y=dep_delay)) + 
  geom_bar(stat="identity", color="black", 
           position=position_dodge()) +
  geom_errorbar(aes(ymin=dep_delay-sd, ymax=dep_delay+sd), width=.2,
                 position=position_dodge(.9)) +
  geom_text(aes(label = round(dep_delay,1)), vjust = 1.5, colour = "white") +
  xlab("Plane Age Group") + 
  ylab("Average Departure Delay (Min)") + 
  ggtitle("The Age Group vs. Departure Delay with Error Bars")
```

