---
title: "STAT 433 - Project"
author: "Junhyeok Park"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
```

## Pre-Processing

First few rows of the raw data are as follows:

```{r echo=FALSE, message=FALSE}
# read in the data
df_p90_p10 = read_csv("./data/p90-vs-p10-logs.csv")

# rename the columns
df_p90_p10 = df_p90_p10 %>% rename(
    "country" = "Entity",
    "code" = "Code",
    "year" = "Year",
    "p90" = "P90",
    "p10" = "P10",
    "population" = "Population (historical estimates)",
    "continent" = "Continent"
  )

df_p90_p10 %>% head(10)
```

### Description on Data Measures

- `country`: the name of the country
- `code`: country code (3 capital letters)
- `year`: the year in which the income distribution was recorded
- `p90`: 90th income percentile
- `p10`: 10th income percentile
- `population`: estimated population of the country
- `continent`: the continent to which the country belongs

```{r include=FALSE}
# extract continents
continent_df = df_p90_p10 %>%
  filter(!is.na(continent)) %>%
  select(country, continent)

# for each country, get income values between the earliest and the lastest avaliable data
# after filling missing continent values
df = df_p90_p10 %>%
    select(-continent) %>% 
    left_join(continent_df, by = "country") %>% 
    filter(!is.na(p90) | !is.na(p10)) %>%
    group_by(country) %>%
    slice(which(year >= min(year) & year <= max(year)))
```

Few rows of the cleaned data are as follows:

```{r echo=FALSE}
df %>%
  slice_max(order_by = p90, n=1) %>%
  head(10)
```

One of our final visualization will look something like this. 

```{r echo=FALSE}
# axes are log transformed
df %>%
  ggplot(aes(x=p10, y=p90, color = continent)) + 
  geom_point() + 
  geom_abline(intercept = 0, slope = 1, color = "gray", lty = "dashed") +
  annotate('text', x=4.5, y=3, label = "1:1", color = "gray") + 
  geom_abline(intercept = 1, slope = 1, color = "gray", lty = "dashed") +
  annotate('text', x=5.5, y=100, label = "10:1", color = "gray") + 
  scale_x_log10(labels=scales::dollar_format()) + 
  scale_y_log10(labels=scales::dollar_format()) + 
  xlab("P10 Income Distribution") + 
  ylab("P90 Income Distribution") +
  labs(color = "Continent") 
```

### Further Analysis

1. Income inequality based on the continents. 
2. Any specific countries? (US, Canada, China, Europe, etc)
3. Any other interesting data to consider?
  - [gdp per capita vs income inequality](https://ourworldindata.org/grapher/gdp-per-capita-vs-economic-inequality)
  - [income inequality](https://ourworldindata.org/grapher/economic-inequality-gini-index)
  - [share of population living in extreme poverty](https://ourworldindata.org/explorers/poverty-explorer)

```{r include=FALSE}
knitr::knit_exit()
```

## Based on the Continents
```{r}
df %>%
  filter(country == "United States" | country == "China") %>%
  ggplot(aes(x=log(p10), y=log(p90), color = country)) + 
  geom_line() 
```


